---
title: "Fitting TabPFN models in R using reticulate"
author: "Måns Thulin"
format: html
---

## Introduction
TabPFN is a foundational model for tabular data, which can be used instead of models like XGBoost and random forest for predictive modelling using tabular data. It was recently [introduced in a paper in Nature](https://www.nature.com/articles/s41586-024-08328-6) by Hollmann et al. The results in that paper indicate that it can yield better results than current state-of-the-art machine learning models in a wide range of settings.

TabPFN is currently only [available for Python](https://github.com/PriorLabs/tabpfn). An R package is reportedly [in the works](https://priorlabs.ai/docs/). However, using the [`reticulate`](https://rstudio.github.io/reticulate/) package, we can already run it from R. The aim of this short tutorial is to show how this can be done, and how to reformat R data to work with TabPFN.

This tutorial uses the packages `reticulate`, `dplyr`, and `ggplot2`. In addition `palmerpenguins` and `modeldata` are used for datasets. Some familiarity with R and machine learning is assumed (have a look at my book [Modern Statistics with R](https://www.modernstatisticswithr.com/) in case you need to brush up on your skills!).

## Setting up TabPFN for use with R
First, install the [`reticulate`](https://rstudio.github.io/reticulate/) package and [Python](https://www.python.org/).

In a terminal, create a virtual environment (venv), activate it, and install TabPFN:

```{bash}
#| eval: false
python -m venv venv
source venv/bin/activate # On Windows: venv\Scripts\activate
pip install tabpfn
```

Next, download the two `.py` scripts from [the `TabPFN_in_R` GitHub repo](https://github.com/mthulin/TabPFN_in_R).

That's it - we're now ready to go! To get started, open R, load `reticulate` and run the two Python scripts, which make classifiers and regression models based on TabPFN available in R.

Make sure to change the `use_virtualenv()` line below so that the path points to the folder where you created your venv.

```{r}
library(reticulate)
use_virtualenv("~/R/venv")

# Source the Python code for training TabPFN models:
source_python("train_classifier.py")
source_python("train_regression.py")
```


# Example 1: binary classification
Here, we'll use [wine data](http://archive.ics.uci.edu/ml/datasets/Wine+Quality) from Cortez et al. (2009), available from the UCI Machine Learning Repository.

```{r}
#| message: false
#| warning: false
library(dplyr)
white <- read.csv("https://tinyurl.com/winedata1",
                  sep = ";") |> mutate(type = "white")
red <- read.csv("https://tinyurl.com/winedata2",
                sep = ";") |> mutate(type = "red")
white |> bind_rows(red) -> wine
```

There are two types of wine: white and red. The aim is to classify the colour of the wine based on its chemical properties.

We split the data into two data frames, one with predictors (`X`) and one with labels (`y`):

```{r}
X <- wine |> select(-type)
y <- wine |> select(type)
```

We can now train and evaluate the model. We specify that `"white"` is the positive class (this is required for computing ROC curves):

```{r}
model <- train_classifier(X, y, pos_label = "white")
```

The evaluation metrics are:
```{r}
model$auc
model$accuracy
```

The `model` object also contains FPR and TPR values that we can use to plot an ROC curve as follows:

```{r}
library(ggplot2)
roc_data <- data.frame(fpr = model$roc[1][[1]], tpr = model$roc[2][[1]])
ggplot(roc_data, aes(fpr, tpr)) +
    geom_line() +
    geom_line(data = data.frame(fpr = c(0, 1), tpr = c(0, 1)), colour = "red") +
    labs(x = "False positive rate",
           y = "True positive rate",
           title = paste("AUC:", round(model$auc, 3)))
```


By default, TabPFN uses 33 % of the data for training. You can change this using the `test_size` parameter:

```{r}
model <- train_classifier(X, y,
                          pos_label = "white",
                          test_size = 0.66)
```

# Example 2: multiclass classification
We'll now use the Palmer penguins data from Data from Gorman et al. (2014), available in the [palmerpenguins](https://cran.r-project.org/web/packages/palmerpenguins/) package.

TabPFN require missing values to be denote by `NaN`, and categorical variables to be represented as characters. Missing categorical variables should be represented by an empty string. Let's modify our data to follow those requirements (as strange as they may seem to R users!):

```{r}
data(penguins, package = "palmerpenguins")

penguins |> 
    mutate_if(is.factor, as.character) |> 
    mutate(sex = if_else(is.na(sex), "", sex)) ->
    penguins

penguins[is.na(penguins)] <- NaN
```

Next, we split the data into predictors and outcomes and train the model:

```{r}
X <- penguins |>
        select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g, sex) |> 
        mutate(sex = as.character(sex))
y <- penguins |>
        select(species) |>
        mutate(species = as.character(species))

model <- train_classifier(X, y)
```

The accuracy is pretty good:
```{r}
model$accuracy
```


# Example 3: regression
To illustrate how to fit a regression model, we'll use the Ames housing data from Data from De Cock (2011), available in the [modeldata](https://cran.r-project.org/web/packages/modeldata/index.html) package.

As in the previous example, we need to convert factors to characters:

```{r}
data(ames, package = "modeldata")

ames |> 
    mutate_if(is.factor, as.character) ->
    ames
```

We can then split the data and train the model:

```{r}
X <- ames |> select(-Sale_Price)
y <- ames |> select(Sale_Price)

model <- train_regression(X, y)
```

Finally, we can review the RMSE, MAE and R2 of the model:

```{r}
model$RMSE
model$MAE
model$R2
```

That's all. Hopefully we can look forward to a native R package and integration with frameworks like `tidymodels` in the future. Meanwhile, I hope that you have fun playing around with the examples above. Feel free to propose changes in the [GitHub repo](https://github.com/mthulin/TabPFN_in_R).